Article Title

### 2013-08-16 sections/Theory.tex
updated figure3

\subsection{Basic concepts}
Let $x_i \in \mathcal X, i=1\ldots,n$ denote a sample of consecutive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is denoted by the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
For example, in our applicaiton, the firing rate is estimated from somatic calcium fluorescence signals with the mean stimulus response subtracted, making it possible for elements of $x_i$ to take on negative values.

@@

But the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-c} \sum\limits_{i=1}^n (x_i -\hat\mu) (x_i - \hat\mu)^\T
\quad\mbox{where}\;
\hat\mu = \frac 1 n \sum\limits_{i=1}^n x_i
\end{equation}
where $c$ is the sum of temporal correlations \TODO{there must be a better description of this}. For independent observations, $c=1$;  $c>1$ when nearby samples are correlated; the value of $c$ can be estimated from the data.

@@

A particularly useful loss function is the mean squared error (MSE), which is proportional to the square of the Frobenius  norm $\|\cdot\|_F$ of the difference between the matrices: 
\begin{equation}
\mathcal L_e(\hat\Sigma,\Sigma) =\frac 1 p \|\hat\Sigma-\Sigma\|_F^2 = \frac 1 p \Tr\left((\hat \Sigma-\Sigma)(\hat\Sigma-\Sigma)^\T\right)
\end{equation}
The MSE is its own excess loss: $\ell_e(\hat\Sigma,\Sigma) \equiv \mathcal L_e(\hat\Sigma,\Sigma)$.

The Gaussian loss function $\mathcal L_g$  arises from the theory of multivariate normal distributions. When observations are identically and independently distributed according to a multivariate normal distribution with zero means, the log likelihood of the covariance matrix $\Sigma$ with $\hat\Sigma = \frac 1 n \sum\limits_{i=1}^n x_i x_i^\T$ is  
\begin{equation}
L\left(\Sigma \mid \hat\Sigma\right) = -\frac n 2 \ln(2\pi) - \frac n 2 \ln \det \Sigma - \frac n 2 \Tr(\Sigma^{-1} \hat \Sigma)
\end{equation}
Then $\mathcal L_g$ is constructed by rescaling $L\left(\Sigma \mid \hat\Sigma\right)$ and dropping the constant term:
\begin{equation}
\mathcal L_g(\hat\Sigma,\Sigma) 
=  -\frac 2 {pn} L\left(\Sigma \mid \hat\Sigma \right) - \frac 1 p \ln(2\pi) 
\equiv  \frac 1 p\left(\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}) \right) 
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat\Sigma,\Sigma) = \mathcal L_g(\hat\Sigma,\Sigma) - \mathcal L_g(\Sigma,\Sigma)  
= \frac 1 p \left(-\ln \det (\hat \Sigma^{-1} \Sigma) + \Tr(\hat \Sigma^{-1}\Sigma)\right) - 1
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

@@

Under the MSE loss function $\mathcal L_e$, the decomposition is the simple sum $r = b^2 + \varepsilon^2$. 
Under other loss functions, $r$ is an increasing function of both $b^2$ and $\varepsilon^2$ although not generally a simple sum.  
\TODO{As far as I know, the terms `bias' and `variance' are only applicable under a quadratic loss function  like $\mathcal L_e$.  Under other loss functions, we may need to use approximation/estimation error.}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figures/Figure2.pdf}
\caption{
Graphical models corresponding to the low-dimensional targets of the four regularization schemes used in the paper.
\textbf{A}: A diagonal matrix corresponds to a Gaussian graphical model with no dependencies. 
\textbf{B}: In factor analysis, observed nodes are assumed to be influenced by several latent units (``factors") but are otherwise independent. 
\textbf{C}: In the Gaussian graphical model (also known as the Gaussian Markov Field), correlations arise from sparse pairwise linear interactions between visible units. 
\textbf{D}: In the Gaussian graphical model with latent units, correlations arise  between pairs of nodes 
}\label{fig:02}
\end{figure}

\subsection{Estimator A: Shrinkage toward a diagonal}
The most popular covariance regularization schemes use diagonal target estimates and linear shrinkage toward the target.  For example, the target could be the identity matrix:  
\begin{equation}
\hat T = \hat v I
\end{equation}
where $\hat v = \frac 1 p \sum\limits_{i=1}^p(\hat\Sigma_0)_{ii}$ is the mean sample variance.

@@

Finally, the target could be a linear mixture of the common variance and independent variance targets
\begin{equation}
\hat T_\eta = (1-\eta)(\hat\Sigma_0 \circ I) + \eta\hat v I
\end{equation}

The regularized estimator is the linear mixture of $\hat\Sigma_0$ and $\hat T_\eta$:
\begin{equation}
\hat\Sigma_{\eta,\lambda} = (1-\lambda)\hat\Sigma_0 + \lambda \hat T_\eta 
\end{equation}

\subsection{Estimator B: Shrinkage toward a factor model}

\subsection{Estimator C: Sparse partial correlations}

\subsection{Estimator D: Sparse partial correlations with latent units}

A good target estimate is one that has low dimensionality to reduce its variance while still allowing to accurately match typical covariance matrices in the given domain.

A regularized estimate requires the following three choices: 
\begin{enumerate}[  1. ]
\item define a set of low-dimensional target estimates
\item given the data, select the optimal target estimate from the set 
\item given the data, select the optimal amount of biasing toward the target estimate
\end{enumerate}

$\hat\Sigma_0$ is unbiased but, with finite sample sizes, not as close to $\Sigma$ as possible.  More \emph{efficient} estimators can be constructed by \emph{regularization}, which is the deliberate biasing of the usual estmate toward a less variable \emph{target estimate} $\hat T$. Target estimates are low-dimensional approximations cannot represent an arbitrary $\Sigma$, but they also have lower random variability.
Regularization reduces the overall error by striking a favorable balance between the variance  (\emph{estimation error}) and the bias \emph{(approximation error)} of the estimate.

Regularization is the game of minimizing risk $r_\gamma$ by reducing estimation error $\varepsilon_\gamma$ at the cost of increased bias $b_\gamma$, the \emph{bias-variance tradeoff}.  
We follow the convention that $\hat\Sigma_{\gamma=0}$ is the unbiased estimator $\hat\Sigma_0$, \emph{i.e.}\;$b_{\gamma=0}^2 = 0$. 
We define a sequence of approximations $\hat\Sigma_\gamma$ that become less variable as as $\gamma$ increases, which is usually accompanies by increased bias. Then $\hat\Sigma_\gamma$ defines the \emph{regularization path}, for which \emph{regularization intensity} $\gamma$ specifies the location along the path, which may be real-valued or discrete, or even a vector to steer the bias toward multiple targets.

\subsection{Regularization of covariance estimates}
The task of designing a regularized estimators consists of defining a sequence of approximations $\hat\Sigma_\gamma$, called \emph{regularization path} and the optimal location on that path, \emph{regularization intensity} $\hat\gamma$ as  functions of the empirical distribution.

There are two basic approaches to design a family of approximations of a parameter that reduce the estimation error: \emph{dimensionality reduction} and \emph{shrinkage}.

Dimensionality reduction projects the estimate onto a low-dimensional manifold (subspace) within the $p(p+1)/2$-dimensional space of covariance matrices $\Theta$.  The variance in the projection is likely to be lower (if the manifold is smooth).  For example, factor analysis with $\nu$ factors projects (w.r.t.\;$\mathcal L_g$) $\hat\Sigma_0$ onto the low-dimensional manifold of covariance matrices representable by factor models. Thus factor analysis can be used as a regularized covariance estimator \citep{Fan:2008} with $\gamma$ controlling the number of factors (or, alternatively, the nuclear norm).  Another dimensionality reduction technique is \emph{covariance selection} \citep{Dempster:1972}, wherein a fraction of inverse covariance coefficients are constrained to zero and the closest (w.r.t.\;$\mathcal L_g$) solution to $\hat\Sigma_0$ is found.

Covariance shrinkage \citep{Schafer:2005,Ledoit:2004} uses a fixed, low-dimensional \emph{target estimate} $T$ and constructs the estimator as the linear mixture, with $\gamma \in [0,1]$,
\begin{equation}
\hat\Sigma_\gamma = \hat\Sigma_0  + \gamma (T-\hat\Sigma_0) 
\end{equation}
 The target can be a diagonal matrix with sample variances, an identity matrix scaled to the mean variance, a single factor model, or any other low-dimensional estimate.

An effective hybrid of dimensionality reduction and shrinkage can be obtained by using $L_1$-regularization techniques.  For example, graphical lasso \citep{Meinshausen:2006,Friedman:2008}, shrinks the inverse of the covariance toward zero as 
\begin{equation}
\hat\Sigma_\gamma = \arg\min\limits_S \mathcal L_g(S,\hat \Sigma_0) + \gamma \| S^{-1} \|_1
\end{equation}
This approach results in forcing many of the coefficients in the inverse of the estimate to zero, or ``sparsification'' of the the inverse.

Many regularization schemes combine multiple targets or multiple approaches, in which case $\gamma$ can become a vector of regularization parameters \citep{Schafer:2005,Fan:2011,Ma:2013} to be optimized in parallel.

\paragraph{Optimization of the regularization parameter:}
Analytic solutions when exist \citep{Ledoit:2004,Schafer:2005}, nested cross-validation otherwise.

\paragraph{Interpretation of revealed structure:}
Sparseness (dimensionality reduction) are key, but must be supported by data, not emph{ad hoc}. \citep{Fan:2006,Malmersjo:2013}.

>>>>

\subsection{Basic concepts}
Let $x_i \in \mathcal X, i=1\ldots,n$ denote a sample of consecutive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is denoted by the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
For example, in our applicaiton, the firing rate is examplestimated from somatic calcium fluorescence signals with the mean stimulus response subtracted, making it possible for elements of $x_i$ to take on negative values.

@@

But the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-c} \sum\limits_{i=1}^n (x_i -\hat\mu) (x_i - \hat\mu)^\T
\quad\mbox{where}\;
\hat\mu = \frac 1 n \sum\limits_{i=1}^n x_i
\end{equation}
where $c$ is the sum of temporal correlations \TODO{there must be a better description of this}. For independent observations, $c=1$;  $c>1$ when nearby samples are correlated; the value of $c$ can be estimated from the data.
\TODO{Under the Gaussian loss, $\mathcal L_g$ (\autoref{eq:GaussLoss}) the scaling is irrelevant, so we will not spend any time talking about the scaling.  But it would be important if we used $\mathcal L_e$.}

@@

A particularly useful loss function is the mean squared error (MSE), which is proportional to the square of the Frobenius  norm $\|\cdot\|_F$ of the difference between the matrices: 
\begin{equation}\label{eq:MSE}
\mathcal L_e(\hat\Sigma,\Sigma) =\frac 1 p \|\hat\Sigma-\Sigma\|_F^2 = \frac 1 p \Tr\left((\hat \Sigma-\Sigma)(\hat\Sigma-\Sigma)^\T\right)
\end{equation}
The MSE is its own excess loss: $\ell_e(\hat\Sigma,\Sigma) \equiv \mathcal L_e(\hat\Sigma,\Sigma)$.

The Gaussian loss function $\mathcal L_g$  arises from the theory of multivariate normal distributions. When observations are identically and independently distributed according to a multivariate normal distribution with zero means, the log likelihood of the covariance matrix $\Sigma$ with $\hat\Sigma = \frac 1 n \sum\limits_{i=1}^n x_i x_i^\T$ is  
\begin{equation}
L\left(\Sigma \mid \hat\Sigma\right) = -\frac n 2 \ln(2\pi) - \frac n 2 \ln \det \Sigma - \frac n 2 \Tr(\Sigma^{-1} \hat \Sigma)
\end{equation}
Then $\mathcal L_g$ is constructed by rescaling $L\left(\Sigma \mid \hat\Sigma\right)$ and dropping the constant term:
\begin{equation}\label{eq:GaussLoss}
\mathcal L_g(\hat\Sigma,\Sigma) 
=  -\frac 2 {pn} L\left(\Sigma \mid \hat\Sigma \right) - \frac 1 p \ln(2\pi) 
\equiv  \frac 1 p\left(\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}) \right) 
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat\Sigma,\Sigma) = \mathcal L_g(\hat\Sigma,\Sigma) - \mathcal L_g(\Sigma,\Sigma)  
= \frac 1 p \left(-\ln \det (\hat \Sigma^{-1} \Sigma) + \Tr(\hat \Sigma^{-1}\Sigma)\right) - 1
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

@@

Under the MSE loss function $\mathcal L_e$ (\autoref{eq:MSE}), the decomposition is the simple sum $r = b^2 + \varepsilon^2$. 
Under other loss functions, $r$ is an increasing function of both $b^2$ and $\varepsilon^2$ although not generally a simple sum.  
\TODO{As far as I know, the terms `bias' and `variance' are only applicable under a quadratic loss function  like $\mathcal L_e$.  Under other loss functions, we may need to use approximation/estimation error.}

@@

\begin{figure}[htp]
\centering
\includegraphics[width=0.5\textwidth]{figures/Figure2.pdf}
\caption{
Graphical models corresponding to the low-dimensional targets of the four regularization schemes used in the paper.
\textbf{A}: A diagonal matrix corresponds to a Gaussian graphical model with no dependencies. 
\textbf{B}: In factor analysis, observed nodes are assumed to be influenced by several latent units (``factors") but are otherwise independent. 
\textbf{C}: In the Gaussian graphical model (also known as the Gaussian Markov Field), correlations arise from sparse pairwise linear interactions between visible units. 
\textbf{D}: In the Gaussian graphical model with latent units, correlations arise  between pairs of nodes 
}\label{fig:02}
\end{figure}

\subsection{Estimator A: Shrinkage toward diagonal}
The most popular covariance regularization schemes use diagonal target estimates and linear shrinkage toward the target.  For example, the target could be the identity matrix:  
\begin{equation}
\hat T = \hat v I
\end{equation}
where $\hat v = \frac 1 p \sum\limits_{i=1}^p(\hat\Sigma_0)_{ii}$ is the mean sample variance.

@@

Finally, the target could be a linear mixture of the common variance and independent variance targets
\begin{equation}
\hat T_{\hat\eta} = (1-\hat\eta)(\hat\Sigma_0 \circ I) + \hat\eta\hat v I
\end{equation}

The regularized estimator is the linear mixture of $\hat\Sigma_0$ and $\hat T_{\hat\eta}$:
\begin{equation}
\hat\Sigma_{\hat\eta,\hat\lambda} = (1-\hat\lambda)\hat\Sigma_0 + \hat\lambda \hat T_{\hat\eta} 
\end{equation}

The hyperparameters $\hat\eta$ and $\hat\lambda$ must be estimated from the data.  Under the MSE loss function $\mathcal L_e$ (\autoref{eq:MSE}), the optimal values of $\hat\eta$ and $\hat\lambda$ can be estimated analytically \citep{Ledoit:2004,Schafer:2005,Schaefer:2010}. However, these estimates are no longer optimal under the Guassian loss $\mathcal L_g$. \TODO{I have tested this and there was a substantial difference in both synthetic and empirical data}  When an analytical solution for optimal hyperparameters is not available, the optimal values can be found by cross validation, entirely within the the training sample. \TODO{explain nested cross-validation in more detail?}

\TODO{Describe how the empirical loss is convex in the hyperparameters in this case}

\subsection{Estimator B: Shrinkage toward a factor model}
In estimator B, the target is a factor model, composed as the sum of a low-rank component $\hat L_{\hat d}\hat L_{\hat d}^\T$ and a diagonal matrix $\hat \Psi$:
\begin{equation}
\hat T_{\hat d} = \hat L_{\hat d} \hat L_{\hat d}^\T + \hat\Psi
\end{equation}
Here $\hat L_{\hat d}$ is an $p\times\hat d$ matrix and $\hat\Psi$ is diagonal.

If $F$ were assume to be multivariate normal \TODO{or other distributions that are defined by linear dependencies}, the factor model is corresponds to the graphical model (\autoref{fig:02}B) where the activity of all neurons are dependent only $\hat d$ latent units. This suggest a mechanistic interpretation in which the interactions between neurons are insignificant compared to the influence of inputs outside the recorded circuit.

Just as in Estimator A, Estimator B is allowed to only partially commit to the low-dimensional target: the overall estimate is the linear mixture of the sample covariance and the target
%\begin{equation}
%\hat\Sigma_{\hat d,\hat\lambda} = (1-\hat\lambda)\hat\Sigma_0 + \hat  
%\end{equation}

\TODO{Check \citep{Fan:2011,Fan:2006}}

\subsection{Estimator C: Sparse partial correlations}
\citep{Meinshausen:2006,Friedman:2008}

\subsection{Estimator D: Sparse partial correlations with latent units}
\citep{Ma:2013}
